Spike spectra for recurrences

### 2022-11-04 Manuscript/inter_spike_manuscript_mdpi.tex

Recurrence plots (RPs) provide a vivid representation of complex dynamics $\vec{x}_i$ stemming from potentially high dimensional systems \cite{marwan2007}
\begin{equation}\label{eq_rp_definition}
R_{i,j}(\varepsilon) = \Theta\left(\varepsilon - D_{i,j}\right) 
= \Theta\left(\varepsilon - \| \vec{x}_i - \vec{x}_j\|\right), \qquad \vec{x} \in \mathbb{R}^d, \quad i,j \in [1,\ldots, N],
\end{equation}
with $\mathbf{R}$ the recurrence matrix, satisfactory $\vec{x}_i$ the state vector at time 
$t = \Delta t \cdot i$ ($\Delta t$ the sampling time), $N$ the number of
sampling points (or length of data series), and $d$ the dimension of the system.
The crucial free parameter $\varepsilon$ is the recurrence threshold, determining what is a recurrences
and, thus, the visible structures in the RP. It can be chosen such that the recurrence rate 
$RR(\varepsilon)=N^{-2}\sum_{i,j}^N R_{i,j}(\varepsilon)$ exhibits a certain value \cite{kraemer2018}.
The simple idea to track recurring states of the $d$-dimensional trajectory $\vec{x}_i$ of the system under study not only allows for a beneficial visualization of the dynamics, 
but also for its 
quantification, using certain structures in the RP, such as diagonal or vertical lines \cite{marwan2007}.

@@

\citet{Zbilut2008} pointed out that $\tau$-RR could be interpreted as analogous to the auto-correlation function $C(\tau)$ for a scalar time series.
However, as there is no equivalent of the Wiener-Khinchin theorem its Fourier Transform (FT) only provides a ``generalized'' kind of power spectral density.
There are clear advantages to such a recurrence-derived power spectral density, the (FT) $\tau$-RR (Fig.~\ref{fig_tau_rr_spectrum_example}D,E):
There are no assumptions for stationarity or sampling, when constructing a RP, 
and since an RP can represent high dimensional dynamics the approch directly applies to higher dimensional systems for which the auto-correlation is a \todo{hm, I'm not so happy with the "matrix" here} matrix and no single power spectrum captures the recurrences.

\todo[inline]{I agree that the following is not necessary and could be removed}
\textbf{Please check the above paragraph for correctness, I did not understand the following and would drop it (or explain the point better):}
\textit{This is reasonable, since the average distances for a given lag $\tau$ 
\begin{equation}
\overline{D}(\tau) = \frac{1}{N-\tau}\sum_{i=1}^{N-\tau} D_{i, i+\tau}
\end{equation}
can be directly read from the distance matrix $\mathbf{D}$ and is also preserved in its thresholded version $\tau$-RR.}

@@

However, the exact meaning of this generalized power spectral density is unclear, and it is typically hard to interpret. $\tau$-RR spikes already directly encode the periodicity of the underlying signal, in contrast it is unclear what, if any, interpretation the ``power'' contained in a particular frequency mode of the $\tau$-RR should have. \textbf{In particular it does not imply anything about the power associated to certain frequencies in the underlying timeseries.} Whenever $\tau$-RR is a spike-train-like signal, which it is in most cases (see Fig.~\ref{fig_tau_rr_spectrum_example}) especially for 
map-data (low-resolution data), a FT of such a signal leads to a spike-train-like image in the frequency domain (e.g., \cite{Schild1982,Cordoba1989}, see 
Fig.~\ref{fig_tau_rr_spectrum_example}E). Thus, it is not intuitive how to extract meaningful information about the systems' state space trajectory.

@@

In this article we propose a new way of decomposing a spike-train-like signal into periodic components as 
an alternative to the RP-based method suggested in \cite{Zbilut2008} or Fourier-based spike-train power 
spectra \cite{Dummer2014}. We instead investigate decomposing the $\tau$-RR in an overcomplete basis of spike trains using a sparsity regularization. Section \ref{sec_tau_rr_method} introduces the approach in detail. Our novel \textit{inter-spike spectrum} does not show resonance 
behavior of the signal's inherent inter-spike frequencies (Fig.~\ref{fig_tau_rr_dirac_comb}C, F). This approach can be used to decompose arbitrary signals, and is not specific to $\tau$-RR. The more spiky the signal is, the more useful our new decomposition can potentially be. In Section \ref{sec_tau_rr_application} we demonstrate the usage of inter-spike spectrum when transforming the $\tau$-RR of a system under study. In this case, the inter-spike spectrum 
can unravel characteristic time scales of high dimensional systems, which is not possible when using a FT. Finally, in Section \ref{sec_tau_rr_conclusion} our results are summarized.

@@

To obtain the Inter-Spike Spectrum , the signal, which we would like to analyze, in our case the $\tau$-RR, is decomposed into a set of appropriate basis functions. 
The general idea common to many methods is that the sum of these weighted functions can approximate a finite signal to a sufficient degree. The weights (in some contexts also called modes or loadings) corresponding to the individual basis functions must be determined.
A number of decomposition techniques based on different sets of basis functions exist, e.g., 
trigonometric functions (Fourier- and wavelet analysis \cite{Bracewell1986}), 
eigenvectors of the corresponding covariance matrix (principal component analysis \cite{Hotelling1933} and related techniques) 
or intrinsic mode functions (empirical mode decomposition and Hilbert spectrum \cite{Norden1998}).
These methods typically share the property that the basis functions form a complete basis,
the set of basis functions is linearly independent, and thus the weights are uniquely determined.

Here we propose to use Dirac combs (DC) with 
different inter-spike periods as basis functions, Eq.~\eqref{eq_dirac_comb}. Let $s(t_i)$ be the min-max-normalized signal we want to transform of length $N$ and 
$t_i=i\cdot \Delta t,~i=1,\ldots,N$, where $\Delta t$ denotes the sampling time and $s(t_i) \in [0,\ 1]\ \forall\ i$. In the following we label this time series as a $(1\times N)$-dimensional 
vector $\vec{s}$. 
First, $\tilde{N}$ different DCs of length $N$ are constructed with inter-spike periods $T_\text{is} \in [1,\ldots,\tilde{N}]$ and $\tilde{N}=\ceil*{N/2}+1$. Second, in order to account 
for possible phase shifts of 
these basis functions occurring in $\vec{s}$, each of these $\tilde{N}$ different DCs also need to be shifted one step further $T_\text{is}-1$ times. This leaves us with a total number of 
$M = \sum_{i=1}^{\tilde{N}}i$ 
basis functions which we can arrange as rows of a $(M\times N)$-sized matrix $\bfX$
(Fig.~\ref{fig_tau_rr_basis_functions} illustrates the described procedure):
\begin{align}
\label{eq_basis_matrix} 
\bfX_{i,j} &= \sum_{k=0}^{N}\delta\bigl(j-1-kT_\text{is}(i)-i+T_\text{is}(i) \bigr), \quad i=1,\dots,\ceil*{N/2}+1, \quad j=1,\dots,N\\
\label{eq_T_is} T_\text{is}(i)&= n, \quad \text{such that}\quad n:~ \frac{n(n-1)}{2} + 1 \leq i <  \frac{n(n+1)}{2} + 1,\quad n \in \mathbb{N}_+.
\end{align}

@@

The Fourier frequency spectra for the Central Europe data in Figure~\ref{fig_power_grid_spectra}A, however, does not display sharp peaks exactly at $15$, $30$ and $60$ \si{min} 
(for technical details on the calculation of 
the spectra shown, the reader is referred to the Appendix~\ref{sec_power_grid_appendix}). Especially the $30$ and $60$ \si{min} 
peaks are split into two adjacent peaks and the local minimum in between these ``double''-peaks correspond to the exact times. The Great Britain counterpart in Fig.~\ref{fig_power_grid_spectra}B 
has sharp peaks at $15$ and $30$ \si{min} and also ``local predecessor peaks'' for $30$ and $60$ \si{min} at the same positions as in panel A. 
In contrast, the $15$ \si{min} peak is completely absent from the inter-spike spectra of the $\tau$-RRs of the frequency data (Fig.~\ref{fig_power_grid_spectra}C,D), but these show sharp peaks 
at $30$ and $60$ \si{min}. In front of these peaks, smaller local peaks can be seen, which correspond to the local peaks of the Fourier power spectra at $28$ and $55$ \si{min}, respectively.
Moreover, there is an additional peak at 
$40$ \si{min} for both datasets, which is absent in the Fourier spectrum and which is not a multiple of the missing $15$ \si{min} peak. The position and magnitude of the peaks in the shown 
inter-spike spectra are robust to the chosen recurrence threshold, the regression method and its regularization as well as the sampling time of the original signal.\\

We interpret the results presented as follows. The missing $15$ \si{min} in the inter-spike spectra is due to the much stronger autocorrelation at $30$ and $60$ \si{min} 
(see Fig.~\ref{fig_power_grid_time_series}B,D) and because these periods are integer multiples of $15$ \si{min} the inter-spike spectra are not able to detect it and the 
sparse regression ``drags'' the $15$ \si{min} periods into the mentioned $30$ and $60$ \si{min} peaks. Moreover, the inter-spike spectra, unlike the Fourier spectra demonstrates sharp peaks exactly 
at $30$ and $60$ \si{min}, i.e. during dispatches (not valid in case of the GB dataset). Eventually, here we have found a clear sharp peak at $40$ \si{min} which can occur 
because of any regular controls in a power grid, and can be a hint to develop the existing stochastic processes to model precisely the power grid frequency \cite{gorjao2020data}.

@@

A novel decomposition technique is proposed that yields the \textit{inter-spike spectrum}. The method decomposes any arbitrary signal into basis functions which consist of (lagged) Dirac 
combs (DC) of different inter-spike period. The weight for each period is obtained by a regularized regression, which promotes sparsity in its solution. We chose LASSO or a sequentially 
thresholded least squares regression STLS in this letter. Since there are 
$M = \sum_{i=1}^{\ceil*{N/2}+1}i$ basis functions for a signal of length $N$ the regression can get computationally intensive for $N>1,000$. When plotting the computed weights as a function 
of the period (or frequency) the inter-spike spectrum is obtained. This provides information that is complementary to the Fourier decompostion of the signal.

Due to the sparse regression underlying the method, there is no unique inverse of the transformation and the regularization parameter plays a crucial role and determines the 
appearance of the obtained inter-spike spectrum. Moreover, similar to the Nyquist frequency barrier in the Fourier Transform which sets a lower bound for the corresponding wave period, here 
the maximum considered inter-spike period is bounded by $T_{\text{is}}^{\text{max}} = \ceil*{N/2}+1$.

Applied to the $\tau$-RR this novel spectrum directly reveals the peridocities of the underlying system, also in cases where there is no direct analogue of the power spectrum.
This general idea \cite{Zbilut2008} allows for a frequency analysis of high dimensional systems, because the RP is a representation of the system's state space trajectory.   
The $\tau$-RR of a recurrence plot (RP) usually has a spiky shape, especially for map-like data, the spikes encode the systems periods, and the inter-spike spectrum can reliably reveal the system's dominant frequencies, 
which is not possible when Fourier transforming the $\tau$-RR or the underlying signal itself. Since the position of the peaks 
in the $\tau$-RR are not sensitive to noise, the corresponding inter-spike spectrum also yields robust results in the presence of noise.

We could think of a broad range of applications of the proposed idea. The inter-spike spectrum itself can serve as a valuable tool for the analysis of any sort of 
spike-train-like data. On the other hand, the inter-spike spectrum of the $\tau$-RR of a signal can serve as a generalized, nonlinear frequency analysis tool for high dimensional complex systems. 
When there is only a subset of state variables available, the state space has to be reconstructed as a pre-processing step. Recent findings \cite{Kraemer2021,Kraemer2022} show that 
this reconstruction process can be reliably automated and applied to multivariate data as well. This would allow for a ``running window'' approach, in order to detect transitions. 
Due to the mentioned computational constraints of our proposed method, a window size $w\leq 1,000$ would possibly suffice for most data, especially when it is map-like, i.e., not 
highly sampled.

>>>>

Recurrence plots (RPs) provide a vivid representation of complex dynamics $\vec{x}_i$ stemming from potentially high dimensional systems \cite{marwan2007}
\begin{equation}\label{eq_rp_definition}
R_{i,j}(\varepsilon) = \Theta\left(\varepsilon - D_{i,j}\right) 
= \Theta\left(\varepsilon - \| \vec{x}_i - \vec{x}_j\|\right), \qquad \vec{x} \in \mathbb{R}^d, \quad i,j \in [1,\ldots, N],
\end{equation}
with $\mathbf{R}$ the recurrence matrix, $\vec{x}_i$ the state vector at time 
$t = \Delta t \cdot i$ ($\Delta t$ the sampling time), $N$ the number of
sampling points (or length of data series), and $d$ the dimension of the system.
The crucial free parameter $\varepsilon$ is the recurrence threshold, determining what is a recurrences
and, thus, the visible structures in the RP. It can be chosen such that the recurrence rate 
$RR(\varepsilon)=N^{-2}\sum_{i,j}^N R_{i,j}(\varepsilon)$ exhibits a certain value \cite{kraemer2018}.
The simple idea to track recurring states of the $d$-dimensional trajectory $\vec{x}_i$ of the system under study not only allows for a beneficial visualization of the dynamics, 
but also for its 
quantification, using certain structures in the RP, such as diagonal or vertical lines \cite{marwan2007}.

@@

\citet{Zbilut2008} pointed out that $\tau$-RR could be interpreted as analogous to the auto-correlation function $C(\tau)$ and, hence, via the Wiener-Khinchin theorem, provide an analogue,
``generalized'', spectral density. This is reasonable, since the average distances for a given lag $\tau$ 
\begin{equation}
\overline{D}(\tau) = \frac{1}{N-\tau}\sum_{i=1}^{N-\tau} D_{i, i+\tau}
\end{equation}
can be directly read from the distance matrix $\mathbf{D}$ and is also preserved in its thresholded version $\tau$-RR. There are clear advantages for a recurrence-derived spectral density, 
i.e., Fourier transforming (FT) $\tau$-RR (Fig.~\ref{fig_tau_rr_spectrum_example}D,E), instead 
of $C(\tau)$: There are no assumptions for stationarity or sampling, when constructing a RP.
Furthermore, since an RP can represent high-dimensional dynamics and its $\tau$-RR serves as a plug-in for $C(\tau)$, the correlation structures of higher dimensional spaces can be read from 
the recurrence-derived Fourier-spectrum.\\

@@

However, the interpretation of this generalized power spectral density is unclear, and it is typically hard to interpret. $\tau$-RR directly encodes the periodicity of the underlying signal, in contrast it is unclear what, if any, interpretation the "power" contained in a particular frequency mode of the $\tau$-RR should have. Whenever $\tau$-RR is a spike-train-like signal, which it is in most cases (see Fig.~\ref{fig_tau_rr_spectrum_example}) especially for 
map-data (low-resolution data), a FT of such a signal leads to a spike-train-like image in the frequency domain (e.g., \cite{Schild1982,Cordoba1989}, see 
Fig.~\ref{fig_tau_rr_spectrum_example}E). Thus, it is not intuitive how to extract meaningful information about the systems' state space trajectory.

@@

In this article we propose a new way of decomposing a spike-train-like signal into periodic components as 
an alternative to the RP-based method suggested in \cite{Zbilut2008} or Fourier-based spike-train power 
spectra \cite{Dummer2014}. This novel \textit{inter-spike spectrum} does not show resonance 
behavior of the signal's inherent inter-spike frequencies in such a way that harmonics of these frequencies are also excited (Fig.~\ref{fig_tau_rr_dirac_comb}C, F). Section \ref{sec_tau_rr_method} explains the idea. Note that this approach can be used to decompose arbitrary signals, and is not specific to $\tau$-RR. However, the more spiky the signal is, the more useful our new approach is compared to a 
FT. In Section \ref{sec_tau_rr_application} we demonstrate the usage of inter-spike spectrum when transforming the $\tau$-RR of a system under study. In this case, the inter-spike spectrum 
can unravel characteristic time scales of high dimensional systems, which is not possible when using a FT. Finally, in Section \ref{sec_tau_rr_conclusion} our results are summarized.

@@

The signal, which we would like to transform\footnote{We focus on $\tau$-RR in this article, Eq.~\eqref{eq_tau_rr}, since 
such time series are often spiky, which favors the proposed decomposition technique. But the proposed method can be applied to any sort of signal.} is decomposed into a set of appropriate basis functions. 
The general idea is that the sum of these weighted functions can approximate a finite signal to a sufficient degree. The weights (also called modes) corresponding to the individual basis functions can also 
be called \textit{loadings} and must be determined. A number of decomposition techniques are out there based on different sets of basis functions, e.g., 
trigonometric functions (Fourier- and wavelet analysis \cite{Bracewell1986}), 
eigenvectors of the corresponding covariance matrix (Principal Component Analysis \cite{Hotelling1933} and related techniques) 
or Intrinsic Mode Functions (Empirical Mode Decomposition and Hilbert spectrum \cite{Norden1998}).

Here we propose to use Dirac combs (DC) with 
different inter-spike periods as basis functions, Eq.~\eqref{eq_dirac_comb}. Let $s(t_i)$ be the min-max-normalized signal we want to transform of length $N$ and 
$t_i=i\cdot \Delta t,~i=1,\ldots,N$, where $\Delta t$ denotes the sampling time and $s(t_i) \in [0,\ 1]\ \forall\ i$. In the following we label this time series as a $(1\times N)$-dimensional 
vector $\vec{s}$. 
First, $\tilde{N}$ different DCs of length $N$ are constructed with inter-spike periods $T_\text{is} \in [1,\ldots,\tilde{N}]$ and $\tilde{N}=\ceil*{N/2}+1$. Second, in order to account 
for possible phase shifts of 
these basis functions occurring in $\vec{s}$, each of these $\tilde{N}$ different DCs also need to be shifted one step further $T_\text{is}-1$ times. This leaves us with a total number of 
$M = \sum_{i=1}^{\tilde{N}}i$ {\color{red}(let's come back to your example in Fig.3, where $N=4$, regarding your explanation $M=\Sigma_i^{\tilde{N}=3} i= 1+2+3=6$. So finally matrix X should be 6x4 size. However, X does not have this size in Fig.3, which is the source of confusion.)}
basis functions which we can arrange as rows of a $(M\times N)$-sized matrix $\bfX$
(Fig.~\ref{fig_tau_rr_basis_functions} illustrates the described procedure) {\color{red}(Moreover, although you claim above that number of rows in X is M in eq. (5) you say $\tilde{N}$, and in Fig. 3 it is four. You should really make it clear and cons)}:
\begin{align}
\label{eq_basis_matrix} 
\bfX_{i,j} &= \sum_{k=0}^{N}\delta\bigl(j-1-kT_\text{is}(i)-i+T_\text{is}(i) \bigr), \quad i=1,\dots,\ceil*{N/2}+1, \quad j=1,\dots,N\\
\label{eq_T_is} T_\text{is}(i)&= n, \quad \text{such that}\quad n:~ \frac{n(n-1)}{2} + 1 \leq i <  \frac{n(n+1)}{2} + 1,\quad n \in \mathbb{N}_+.
\end{align}

@@

The Fourier frequency spectra for the Central Europe data in Figure~\ref{fig_power_grid_spectra}A, however, does not display sharp peaks exactly at $15$, $30$ and $60$ \si{min}, which may partly 
be explained by the leakage effect (for technical details on the calculation of 
the spectra shown, the reader is referred to the Appendix~\ref{sec_power_grid_appendix}). Especially the $30$ and $60$ \si{min} 
peaks are split into two adjacent peaks and the local minimum in between these ``double''-peaks correspond to the exact times. The Great Britain counterpart in Fig.~\ref{fig_power_grid_spectra}B 
has sharp peaks at $15$ and $30$ \si{min} and also ``local predecessor peaks'' for $30$ and $60$ \si{min} at the same positions as in panel A. 
In contrast, the $15$ \si{min} peak is completely absent from the inter-spike spectra of the $\tau$-RRs of the frequency data (Fig.~\ref{fig_power_grid_spectra}C,D), but these show sharp peaks 
at $30$ and $60$ \si{min} (there is no leakage effect due to the proposed decomposition technique). 
In front of these peaks, smaller local peaks can be seen, which correspond to the local peaks of the Fourier power spectra at $28$ and $55$ \si{min}, respectively.
Moreover, there is an additional peak at 
$40$ \si{min} for both datasets, which is absent in the Fourier spectrum and which is not a multiple of the missing $15$ \si{min} peak. The position and magnitude of the peaks in the shown 
inter-spike spectra are robust to the chosen recurrence threshold, the regression method and its regularization as well as the sampling time of the original signal.\\

We interpret the results presented as follows. It should be noted firstly that ISS is not only capable to recognize the well-known periodicity in the power grid frequency because of the dispatching time, but also the ISS demonstrate more sharp peaks at this periodicity. The missing $15$ \si{min} in the inter-spike spectra is due to the much stronger autocorrelation at $30$ and $60$ \si{min} 
(see Fig.~\ref{fig_power_grid_time_series}B,D) and because these periods are integer multiples of $15$ \si{min} the inter-spike spectra are not able to detect it and the 
sparse regression ``drags'' the $15$ \si{min} periods into the mentioned $30$ and $60$ \si{min} peaks. Moreover, the inter-spike spectra, unlike the Fourier spectra demonstrates sharp peaks exactly 
at $30$ and $60$ \si{min}, i.e. during dispatches (not valid in case of the GB dataset). Eventually, here we have found a clear sharp peak at $40$ \si{min} which can occur 
because of any regular controls in a power grid, and can be a hint to develop the existing stochastic processes to model precisely the power grid frequency \cite{gorjao2020data}.

@@

A novel decomposition technique is proposed that yields the \textit{inter-spike spectrum}. The method decomposes any arbitrary signal into basis functions which consist of (lagged) Dirac 
combs (DC) of different inter-spike period. The loading for each period is obtained by a regularized regression, which promotes sparsity in its solution. We chose LASSO or a sequentially 
thresholded least squares regression STLS in this letter. Since there are 
$M = \sum_{i=1}^{\ceil*{N/2}+1}i$ basis functions for a signal of length $N$ the regression can get computationally intensive for $N>1,000$. When plotting the computed loadings as a function 
of the period (or frequency) the inter-spike spectrum is obtained. A disadvantage is that the transformation is not invertible. An advantage is that there is no leakage effect.
Although this novel spectrum is superior to an ordinary FFT-based power spectrum when the signal has a spike-train-like appearance, the authors suggest that this method should 
be considered 
as an additional source of information but not as a substitute for ordinary Fourier analysis. 
Due to the sparse regression underlying the method, there is no unique inverse of the transformation and the regularization parameter plays a crucial role and determines the 
appearance of the obtained inter-spike spectrum. Moreover, similar to the Nyquist frequency barrier in the Fourier Transform which sets a lower bound for the corresponding wave period, here 
the maximum considered inter-spike period is bounded by $T_{\text{is}}^{\text{max}} = \ceil*{N/2}+1$.

The invention of the proposed method has been motivated by the idea of transforming $\tau$-recurrence rate signals ($\tau$-RRs) into their frequency domain. 
This general idea \cite{Zbilut2008} allows for a frequency analysis of high dimensional systems, because the RP is a representation of the system's state space trajectory.   
The $\tau$-RR of a recurrence plot (RP) usually has a spiky shape, especially for map-like data, and the inter-spike spectrum can reliably reveal the system's dominant frequencies, 
which is not possible when Fourier transforming the $\tau$-RR or the underlying signal itself. Since the position of the peaks 
in the $\tau$-RR are not sensitive to noise, the corresponding inter-spike spectrum also yields robust results in the presence of noise.

We could think of a broad range of applications of the proposed idea. The inter-spike spectrum itself can serve as a valuable tool for the analysis of any sort of 
spike-train-like data. On the other hand, the inter-spike spectrum of the $\tau$-RR of a signal can serve as a generalized, nonlinear frequency analysis tool for complex systems. 
When there is only a subset of state variables available, the state space has to be reconstructed as a pre-processing step. Recent findings \cite{Kraemer2021,Kraemer2022} show that 
this reconstruction process can be reliably automated and applied to multivariate data as well. This would allow for a ``running window'' approach, in order to detect transitions. 
Due to the mentioned computational constraints of our proposed method, a window size $w\leq 1,000$ would possibly suffice for most data, especially when it is map-like, i.e., not 
highly sampled.
